---
title: "Cover letter: Making, Updating, and Querying Causal Models with `CausalQueries`"
format: pdf
---

Berlin, `r format(Sys.Date(), "%B %Y")`


Dear Editors:

Thank you and the reviewers for these very useful comments on our manuscript. 


Reviewers 1 provided useful guidance especially regarding coding, formatting, and style.

We address these in turn:

1. `\section`, `\subsection`, etc. should be in sentence style  

> *We have implemented this change.*

2. The code presented in the manuscript should not contain comments within the verbatim code. Instead the comments should be made in the normal LaTeX text.

> *We have removed all comments from verbatim code.*

3. For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the text width (up to 76 or 77 characters) and be indented by two spaces,  

> *We relied on the `.qmd` template for jss; we have now modified this for `CodeInput`; for `CodeOutput` we adjusted the `.tex` files after compiling.*

4. Code should have enough spaces to facilitate reading. Please include spaces before and after operators and after commas (unless spaces have syntactical meaning).  

> *We have checked compliance with this provision.*

5. For R-related manuscripts: The first argument of data() and library() should always be quoted, e.g., library("foo").  

> *We have checked compliance with this provision.*

6. If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".  

> *We have checked compliance with this provision.*

7. The rule for capitalizing the starting letters of Figure, Section and Table is as follows: If you are referring to a particular figure/section/table then capitalize the first letter, otherwise use a lower-case first letter.  

> *We have checked compliance with this provision.*

8. Figures, tables and equations should be marked with a `\label` and referred to by `\ref`  

> *We have checked compliance with this provision.*

9. All captions should appear below the corresponding figure/table. The captions should be in sentence style and end with a period. No additional formatting should be used for the caption.  

> *We have checked compliance with this provision.*

10. All table row/column headers should also be in sentence style. There should not be further footnote-style annotations in tables; these should all be placed in the caption.  

> *We have checked compliance with this provision.*

11. Equations should be marked with a `\label` and referred to by either `Equation~\ref{...}` (with capitalization, without parentheses) or (`\ref({...}`) with the former being preferred if the number of equation references is not too large.  

> *We have checked compliance with this provision.*

12. In all cases, code input/output must fit within the normal text width of the manuscript.  

> *We have checked compliance with this provision.*

13. For bullet lists/itemized lists please use either a comma, semi-colon, or period at the end of each item.  

> *We have checked compliance with this provision.*

14. Abbreviations should be spelled in upper-case letters without additional formatting (i.e., without periods, without small caps, italics, etc.). All abbreviations should be introduced with their expansion where the expansion should not be capitalized.  

> *We have checked compliance with this provision.*

15. Do not use additional formatting for specific words unless explicitly required by the JSS style guide  

> *We have checked compliance with this provision.*

16. As a reminder, please make sure that: `\proglang`, `\pkg` and `\code` have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.  

> *We have checked compliance with this provision.*

17. Please make sure that all software packages are \cite{}'d properly.  

> *We have checked compliance with this provision.*

18. All references should be in title style.  

> *We have checked compliance with this provision.*

# Reviewer 2

Reviewer 2 provided a number of very helpful substantive comments as well as useful suggestions for the package.

1. Page 6, paragraph 4, line 1: "...define arbitrary DAG..." should be "...define an arbitrary DAG...".  

> *Corrected, with thanks.*

2. $2^{(4^2)}$ with $2^{(2^4)}$  

> *Corrected, with thanks.*

3. Page 7, Definition 1, item 3: For consistency with items 1 and 2, the authors may want to replace "a collection of functions..." with "an ordered collection of functions...".  

> *Edited, with thanks.*

4. Also, since the paper's introduction refers to "...causal models consistent with the structural model..." it would be beneficial to add a note on how these terms are distinguished in this paper somewhere.  

> *We add a paragraph distinguishing between these types of model beside the introduction of the definition.*

5. Page 7, first paragraph following Definition 1: For readers familiar with Pearl’s work, the authors may want to add a note that the nodal type variables in this paper often are referred to as response function variables in his terminology.  

> *Edited, with thanks.*

6. Page 7, third paragraph following Definition 1: It is unclear what "...align the domain..." means. Using "...partition the domain..." is more clear and standard terminology. Pearl calls this a canonical partition.  

> *Edited, with thanks.*

7. $f^j \rightarrow f_{Y_j}$, and similar  

> *Edited, with thanks.*

8. Page 8, line 14: Following common terminology on distributions over variables, "...Dirichlet distribution over $\lambda^j$..." should be either "...Dirichlet distribution over $\theta^j$..." or "...Dirichlet distribution for $\lambda$...". Likewise, the displayed Bayesian updating equation following the paragraph should use $\theta$ in place of $\lambda$, since $\lambda$ is itself the probability of the causal type $\theta$.  

> *Please note: We have maintained the existing language here. The reason is that the `CausalQueries` package in fact updates over the $\lambda$s. The reviewer correctly notes that the $\lambda$ terms characterize the distribution over the $\theta$s; they are thus part of the definition of a causal model. However `CausalQueries` works on distributions over the $\lambda$s, that is it lets users update over causal models. Intuitively one might think of $\lambda^j$ as capturing the share of units of type $\theta^j$; if a unit is drawn at random then the probability that a unit has type $\theta^j$ is $\lambda^j$. We assume however that there is uncertainty not just over $\theta^j$ but over the shares, $\lambda^j$ and so we update over these shares.*

9. Page 8, bottom: In the displayed multinomial, the parameter N has not been defined.  

> *Edited, with thanks.*

10. Page 9, first displayed formula: Here it becomes relevant what kinds of responses the subscripts actually encode.  

> *This is correct, thank you. We have now added text providing the encoding we use explicitly: "As a practical matter we need to label response types. In [CausalQueries]{.pkg} this is done using subscripts that indicate the response given different combinations of parents. A node, $Y$, with one binary parent, $X$, has nodal types subscripted with two values indicating the two possible values of $Y$'s parent ($0$ or $1$): $(\theta_{00}, \theta_{01}, \theta_{10}, \theta_{11})$, where $\theta_{ab}$ denotes a nodal type for which $Y$ takes the value $a$ under the intervention $\text{do}(X = 0)$ and the value $b$ under $\text{do}(X = 1)$. The same approach is used for nodes with more (or fewer) nodal types, where the $i$th digit in the subscript corresponds to the value the node takes when the parents take on the $i$th combination of possible parent values (listed in colexicographic binary order given the specified ordering of parents)."*

11. Page 16–17: Regarding caution and sensitivity to priors, the authors may also want to refer to Richardson et al. [2011].  

> *Added, with thanks.*

12. Page 17, line 1: In "...when models are not identified...", the word models should rather be replaced by estimands or queries.  

> *Edited, with thanks.*

13. Page 27, second code block should use `"Y[X=0]==1"` rather than `"Y[X=1]==1"` to be consistent with the preceding paragraph.  

> *Corrected, with thanks.*

14. Page 28, first line of paragraph following second code block: "...in addition..." should be "... and in addition,..." or "...in addition to...".  

> *Corrected, with thanks.*

15. Page 29, example under “Nested queries": The authors may want to add a note that the query `Y[M=M[X=0],X=1]==1` is commonly referred to as a natural direct effect in mediation analysis by Pearl.  

> *Added, with thanks.*

16. Page 32, sentence before code block: "...limits model..." should be "...lipids model...".  

> *Corrected, with thanks.*

17. Page 34: The authors note that risk ratios are currently not implemented but trying `lipids_model |> get_query_types("Y[X=1] / Y[X=0]")` yields output without warnings.  

> *The reviewer is correct. The output in this case indicates possible problems but no warning was provided:*

```{r, eval = FALSE}
>lipids_model |> get_query_types("Y[X=1] / Y[X=0]")
    Z0.X00.Y00 Z1.X00.Y00 Z0.X10.Y00 Z1.X10.Y00 Z0.X01.Y00 Z1.X01.Y00 
           NaN        NaN        NaN        NaN        NaN        NaN  
    Z0.X01.Y10 Z1.X01.Y10 Z0.X11.Y10 Z1.X11.Y10 Z0.X00.Y01 Z1.X00.Y01 
             0          0          0          0        Inf        Inf 
```

> *The output from the updated package now includes a warning:*

```{r, eval = FALSE}
Warning message:
In check_query(query) :
  Non-linear transformations (such as /, ^, exp() or log()) 
  are not supported in querying.
```

18. In `?make_model`, the possibility for confounded `X<->Y` should be mentioned.  

> *Added to the most recent package update, with thanks.*

19. Also nice to have: links to documentations of related functions mentioned, e.g., to set priors and set parameters in `?make_model`.  

> *Implemented in the most recent package update, with thanks.*

20. In `summary.causal_model`, the meaning of the display column for a node is not immediately clear and might benefit from an example in the documentation.

> *We have sought to make this output more interpretable in the most recent package update:*

```{r, eval = FALSE}
> make_model("X1 -> Y <- X2") |> summary() 

Causal statement: 
X1 -> Y; X2 -> Y

Nodal types: 

Nodal types for X1:
0  1

Nodal types for X2:
0  1

Nodal types for Y:
0000  1000  0100  1100  0010  1010  0110  1110  0001  ...

Guide to interpreting nodal types for Y:

  index            interpretation
1  *---  Y = * if X1 = 0 & X2 = 0
2  -*--  Y = * if X1 = 1 & X2 = 0
3  --*-  Y = * if X1 = 0 & X2 = 1
4  ---*  Y = * if X1 = 1 & X2 = 1
```

This careful set of reviews and the close reads have greatly strengthened this manuscript. 

We note only one point of disagreement with reviewer B but hope to have improved the text at this point also to make the logic clearer.

With thanks,  

Macartan Humphreys
